---
title: "Chapter 08 Problem Sets"
format: html
editor: visual
warning: false
---

```{r}
#| echo: false
#| warning: false
#| output: false
library(tidyverse)
library(gt)
library(latex2exp)
library(broom.mixed)
```

## Problem 8.1 Drug trials

Suppose that we are testing the efficacy of a certain drug which aims to cure depression, across
two groups, each of size 10, with varying levels of the underlying condition: mild and severe.
We suppose that the success rate of the drug varies across each of the groups, with $θ_{mild} > θ_{severe}$.
We are comparing this with another group of 10 individuals, which has a success rate equal to
the mean of the other two groups, $θ_{homogeneous} = ( θ_{mild} + θ_{severe} ) / 2$.

## Problem 8.1.1

Calculate the mean number of successful trials in each of the three groups.

::: {style="color:blue"}
Across each of the three groups:

* $\mathbb{E}[X{mild}] = 10\times\theta_{mild}$
* $\mathbb{E}[X{severe}] = 10\times\theta_{severe}$
* $\mathbb{E}[X{homogeneous}] = 10\times\theta_{homogeneous}$

:::

### Problem 8.1.2

Compare the mean across the two heterogeneous groups with that of the single
group of 10 homogeneous people.
::: {style="color:blue"}

Taking the mean of the means for each group,

$$
\begin{aligned}
\mathbb{E}[X{combined}] &= \frac{1}{2}\times10 (\theta_{mild} + \theta_{severe}) = 10\times\theta_{homogeneous}  
\end{aligned}
$$

In words, the mean outcome across the two groups is the same.
:::
## Problem 8.1.3

Calculate the variance of outcomes across each of the three groups.

::: {style="color:blue"}
The variance across each of the three groups is given by:


* $var(X{mild}) = 10\times\theta_{mild}(1-\theta_{mild})$
* $var(X{severe}) = 10\times\theta_{severe}(1-\theta_{severe})$
* $var(X{homogeneous}) = 10\times\theta_{homogeneous}(1-theta_{homogeneous})$

:::

### Problem 8.1.4

How does the variance across both heterogeneous studies compare with that of
a homogeneous group of the same sample size and same mean?

::: {style="color:blue"}
TODO
:::

### Problem 8.1.5

Now consider the extension to a large number of trials, where the depressive
status of each group is unknown to the experimenter, but follows $θ ~ beta( α , β )$. Calculate the
mean value of the beta distribution.


::: {style="color:blue"}
$\frac{\alpha}{\alpha+\beta}$
:::

### Problem 8.1.6

Which combinations of $α$ and $β$ would make the mean the same as that of a
single study with success probability $θ$?


::: {style="color:blue"}
TODO
:::

### Problem 8.1.7

How does the variance change, as the parameters of the beta distribution are
changed, so as to keep the same mean of $θ$?


::: {style="color:blue"}
TODO
:::

### Problem 8.1.8

How does the variance of the number of disease cases compare to that of a single
study with success probability $θ$?


::: {style="color:blue"}
TODO
:::


#### Problem 8.1.9

Under what conditions does the variance in disease cases tend to that from a
binomial distribution?


::: {style="color:blue"}
TODO
:::

### Problem 8.2 Political partying

Suppose that in polls for an upcoming election there are three political parties that individuals
can vote for, denoted by $\{ A , B , C \}$ , respectively.

### Problem 8.2.1

If we assume independence among those individuals who are polled then what
likelihood might we choose?


::: {style="color:blue"}
A multinomial likelihood with probabilities of voting for each party given by $(p_A, p_B, p_C)$.
:::

### Problem 8.2.2

In a sample of 10 individuals we find that the numbers who intend to vote for
each party are $( n_A , n_B , n_C ) = (6,3,1)$ . Derive and calculate the maximum likelihood estimators of
the proportions voting for each party.


::: {style="color:blue"}
$$
\hat{p}_i = \frac{n_i}{n}\text{, where }i \in \{A, B, C\}
$$
```{r}
ni <- c(na=6,nb=3,nc=1)
n <- sum(ni)

ni/n
```

:::

## Problem 8.2.3
Dirichlet
Graph the likelihood in $( p_A , p_B )$ space.

::: {style="color:blue"}
TODO
:::


